Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-07-13 manuscript.rmd
Added Inverse model methods section Other smaller edits

---
title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

@@

- Deep water access (WA~Deep~) is believed to be an important, drought-tolerance mechanism, but under-studied due to lack of data. Vulnerability to embolism ($\Psi_{88, Stem}$) and safety margins ($\Psi_{min} - \Psi_{88, Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. Here we test its role in drought-tolerance strategies of a tropical forest community (Panama) with a novel data-model approach.
- We inversely modelled Effective Rooting Depth (ERD) for 32 large, canopy tree species by linking species growth dynamics (xx trees, 6 censuses, 1990-2015) to VPD, $\Psi_{soil}$ by depth in the whole-soil column, and a novel community-scale model of species vulnerability curves ($K_{Leaf}$ vs. $\Psi_{x}$) linking data on the same for 21 species with community-scale softer traits. We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$). 
- We find that deeper ERD is associated with higher hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margin, but lower mean mortality rate. Deeper ERD species spent less time below critical hydraulic thredholds indicating that ERD buffers drought-stress.
- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for modeling the future of tropical species and forests under climate change.

@@

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size. Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottlneck in testing this mechanism community-wide has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 1% of the World’s 45-53 thousand tree species [@Fan:2017; @Silk:2000]. Further, only a small fraction of those are tropical, even though ~95% of the World’s tree diversity resides in the tropics [@Silk:2000]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic concentration in plant tissue to it's gradient vertically belowground. Such data is available only for ~0.05% of World's tree species [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests. Further, to estimate plant available water in the vertical soil column requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see Johnson, Porporato).

Trait-based plant ecology aims to predict impacts of environmental change on community composition and function using relevant plant traits as currency. To reach this goal it is imperative to establish relationships between plant's environment, traits and demographic outcomes, but this remains to be established for trees' realised water environments and water-access through droughts. Huge efforts have been devoted to linking same-for-all-trees drought indices, above-ground traits and demographic rates. These have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88, Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88, Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity. A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88, Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88, Stem}$). A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88, Stem}$ than other two species limited to shallow-water. Chitra-Tarak et al. (2018) proposed a general model to link realised water availability to demographic outcomes through droughts at tree community-scale. This study found that species accessing deep-water in a tropical dry forest study system died more in a rare, centennial drought compared to shallow-water species. Authors concluded that this was consistent with expectation of deep water usage as an aquisitive strategy relying on stable and reliable water resource at depth, which exhausted in the multi-annual drought resulting in reduced survival, compared to persistent species associated with shallow-water that was relatively more available via infiltration during this drought. However, species estimates of water-access modelled by this study were not corroborated with rooting/water-sourcing depths, or related to hydraulic traits.

\
Here, we obtained daily dynamics of plant water potential ($\Psi_{soil, z}$) in the whole-soil column of a neo-tropical forest (BCI, Panama) with a locally parametarised and calibrated 1-D hydrological water balance model. We then estimated the effective rooting depths for xx large, canopy tree species as the depth ($z$) whose soil water dynamics ($\Psi_{soil, z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015) with a novel inverse model. This model incorporated impact of atmospheric drought (VPD) on growth and new data on species leaf vulnerability curves ($K_{Leaf}$ vs. $\Psi_{x}$), which we scaled-up to community level via scaling relationships with commonly available traits. We evaluated the inversely estimated ERD against exisiting data on plant and soil water isotopes ($\delta^2H$).

@@

## The Study Site
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, including availability of an independent isotopic tracer study on trees’ water-sourcing depths (Meinzer et al.). The 1,560 hectare Barro Colorado Island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD over 1983-2017) and a four-month long dry season from mid-December through mid-April  (≤ 100 mm rain/month). The entire Island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% (Condit et al., 2000). Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 (Condit, 1998; Hubbell et al., 1999) (Anderson-Texeira xx).

\
In the 50 ha plot all stems above 10 mm are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for diameter growth, mortality as well as recruitment of new stems into the 10 mm size class (Hubbell, S.P., Condit, R., and Foster, R.B., 2010; Condit et al., 2012). This inventory represents 321 woody species across the eight censuses. Aside from a 2 ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture (Piperno 1992). The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) (Harms 2001). The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found in Croat (1978), Leigh et al. (1982) and Gentry (1990).

@@

\
To estimate dynamics of water availability by depth over the forest’s root zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). FATES is a state-of-the-art, community-based, open-source model widely used for studying climate-vegetation interactions. FATES is considered a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches using the ecosystem demography approach...... (How ELM talks to FATES)...Detailed descriptions can be found elsewhere (Koven et al., 2020) Fisher et al ).

@@

We measured ($K_{x, Leaf}$ vs. $\Psi_{x, Leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x). We fitted exponential, sigmoidal, logistic and exponential form of equations to species-level data as prescribed by [@GUYOT:2011] and based on AIC criteria chose exponential form as the best-fitting one (Fig. \@ref(fig:obs-plc-curves)), described as

\begin{equation}
K_{Leaf} = y_0 + Ae^{-B\Psi_{Leaf}} + error
(\#eq:vc-exp)
\end{equation}

@@

where $(a_0, a_1, a_2, a_3, b_0, b_1, b_2, b_3)$ are coefficients. $WSG$ & parameter $B$ explained large variation in parameter $A$ (Eq. \@ref(eq:a-mod); Adj. R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig. \@ref(fig:ab-fit)). $WSG$ & $LMA$ explained large variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r signif(glance(k_by_psi.models[[1]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[1]])$p.value, 2)`). We thus sequentially used Eq. (\#eq:b-mod) and (\#eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig. \@ref(fig:comm-plc-curves)).

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{x, Leaf}$ vs. $\\Psi_{x, Leaf}$) at community-scale (", nrow(data.model.AB$sp)," species) based on species data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species. Each line represents a species, color coded by $WSG$.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
```

@@

@@

@@

Figure \@ref(fig:erd-iso)

@@

To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance  ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{TLP}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88, Stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88, Stem}$).

@@

We calculated mortality rates for large trees (>=10 cm DBH) of xx canopy species for each of the seven census interval ($int$) in the 33 year record (1985-2015; $Mort_{int}$) and a species mean mortality rate across those intervals ($Mort_{mean} = \frac{1}{int}\sum_{int=1}^{6} {Mort_{int}}$). To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{80,Leaf}$.

@@

\
In this forest modelled Effective Rooting Depths (ERD) for the 32 large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)). For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{TLP}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88, Stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88, Stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).

```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(unique(erd.data$R2)) ," large, canopy species of Barro Colorado Island. See Table S2 for scientific names of species codes")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
```

@@

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{80,Leaf}$ over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

>>>>

---
title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::pdf_document2: default
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

@@

- Deep water access (WA~Deep~) is believed to be an important, drought-tolerance mechanism, but under-studied due to lack of data. Vulnerability to embolism ($\Psi_{88,~Stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. Here we test its role in drought-tolerance strategies of a tropical forest community (Panama) with a novel data-model approach.
- We inversely modelled Effective Rooting Depth (ERD) for 32 large, canopy tree species by linking species growth dynamics (xx trees, 6 censuses, 1990-2015) to VPD, $\Psi_{x,~z}$ by depth $z$in the whole-soil column, and a novel community-scale model of species vulnerability curves ($K_{Leaf}$ vs. $\Psi_{x,~Leaf}$) linking data on the same for 21 species with community-scale softer traits. We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$). 
- We find that deeper ERD is associated with higher hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margin, but lower mean mortality rate. Deeper ERD species spent less time below critical hydraulic thredholds indicating that ERD buffers drought-stress.
- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for modeling the future of tropical species and forests under climate change.

@@

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size. Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottlneck in testing this mechanism community-wide has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic concentration in plant tissue to it's gradient vertically belowground. Such data is available only for ~0.05% of World's tree species [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests. Further, to estimate plant available water in the vertical soil column requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see [@Johnson:2018fad;@Kumagai:2011;@Ivanov:2012].

Trait-based plant ecology aims to predict impacts of environmental change on community composition and function using relevant plant traits as currency. To reach this goal it is imperative to establish relationships between plant's environment, traits and demographic outcomes, but this remains to be established for trees' realised water environments and water-access through droughts. Huge efforts have been devoted to linking same-for-all-trees drought indices, above-ground traits and demographic rates. These have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity. A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88,~Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$). A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88,~Stem}$ than other two species limited to shallow-water. Chitra-Tarak et al. (2018) proposed a general model to link realised water availability to demographic outcomes through droughts at tree community-scale. This study found that species accessing deep-water in a tropical dry forest study system died more in a rare, centennial drought compared to shallow-water species. Authors concluded that this was consistent with expectation of deep water usage as an aquisitive strategy relying on stable and reliable water resource at depth, which exhausted in the multi-annual drought resulting in reduced survival, compared to persistent species associated with shallow-water that was relatively more available via infiltration during this drought. However, species estimates of water-access modelled by this study were not corroborated with rooting/water-sourcing depths, or related to hydraulic traits.

\
Here, we obtained daily dynamics of plant water potential ($\Psi_{x,~z}$) by depth $z$ in the whole-soil column of a neo-tropical forest (BCI, Panama) with a locally parametarised and calibrated 1-D hydrological water balance model. We then estimated the effective rooting depths for xx large, canopy tree species as the depth ($z$) whose soil water dynamics ($\Psi_{x,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015) with a novel inverse model. This model incorporated impact of atmospheric drought (VPD) on growth and new data on species leaf vulnerability curves ($K_{Leaf}$ vs. $\Psi_{x,~Leaf}$), which we scaled-up to community level via scaling relationships with commonly available traits. We evaluated the inversely estimated ERD against exisiting data on plant and soil water isotopes ($\delta^2H$).

@@

## The Study Site
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, including availability of an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560 hectare Barro Colorado Island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD over 1983-2017) and a four-month long dry season from mid-December through mid-April  (≤ 100 mm rain/month). The entire Island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit:1998395; @Hubbell:1999; @Anderson-Teixeira:2014].

\
In the 50 ha plot all stems above 10 mm are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for diameter growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit:2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2 ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture. The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere.

@@

\
To estimate dynamics of water availability by depth over the forest’s root zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). FATES is a state-of-the-art, community-based, open-source model increasingly used for studying climate-vegetation interactions. FATES is considered a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches using the ecosystem demography approach...... (How ELM talks to FATES)...Detailed descriptions can be found elsewhere [@Koven:2020; @Fisher: ).

@@

We measured ($K_{x,~Leaf}$ vs. $\Psi_{x,~Leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x). We fitted exponential, sigmoidal, logistic forms of equations to species-level data as prescribed by [@Guyot:2011] and based on AIC criteria chose exponential form (Fig. S\@ref(fig:obs-plc-curves)), described as

\begin{equation}
K_{Leaf} = Ae^{-B\Psi_{Leaf}} + error
(\#eq:vc-exp)
\end{equation}

@@

where $(a_0, a_1, a_2, a_3, b_0, b_1, b_2, b_3)$ are coefficients. $WSG$ & parameter $B$ explained large variation in parameter $A$ (Eq. \@ref(eq:a-mod); Adj. R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig. S\@ref(fig:ab-fit)). $WSG$ & $LMA$ explained large variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r signif(glance(k_by_psi.models[[1]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[1]])$p.value, 2)`). We thus sequentially used Eq. (\#eq:b-mod) and (\#eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig. \@ref(fig:comm-plc-curves)).

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{x,~Leaf}$ vs. $\\Psi_{x,~Leaf}$) at community-scale (", nrow(data.model.AB$sp)," species) based on species data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species. Each line represents a species, color coded by $WSG$.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
```

@@

With 
\
\begin{equation}
K_{Leaf,z,n} = Ae^{-B\Psi_{z,n}}
(\#eq:k-soil)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,z,n} + error
(\#eq:M1)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,n} StdVPD_{n} + error
(\#eq:M2)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,z,n} LeafFrac_{doy} + error
(\#eq:M3)
\end{equation}

\begin{equation}
Growth_{t} = \frac{1}{n_{t}}\sum_{day=1}^{n_{t}} StdK_{Leaf,z,n} LeafFrac_{doy}StdVPD_{n} + error
(\#eq:M4)
\end{equation}

@@

Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm DBH) and species whose maximum height was at least 30 m (hereafter, canopy species). We chose only those trees for which diameter (DBH) record was present across all size censuses (1990, 19905, 2000, 2005, 2010, 2015). We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ cmyr^{-1}) for each of the five census intervals ($t$). As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a DBH model of tree growth based on a trees' DBH at the beginning of a census onterval ($GrResid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$). The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ). We then standardised (centered and scaled) an individual tree time series ($GrResid_{i,t}$) and obtained species-level (*n* = 32) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx).

As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) and VPD (kPa; $GPP = f(VPD) + error$;  SI) and used standardised $f(VPD)$ (range 0-1) as $StdVPD$ in Eqs. \@ref(eq:M2)- \@ref(eq:M4).

Leaf fraction...

For each model (\@ref(eq:M2)- \@ref(eq:M4)) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp, model we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD); after retaining species with positive $\rho$ and *R*^2^ >= 0.1.

@@

For each model, species ERD was regressed against species Xylem sap $\delta^2H$ (%~$\circ$~; *n* = 7). We used Xylem sap $\delta^2H$ values given in [@Meinzer:2001] for dry-season (March 1997) as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil. We removed species that are deciduous in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced in March and thus may not comparable to other species with leaves present. Fig. S\@ref(fig:iso-data). We also removed *Guapira* *spp*, whose xylem sap value fell in the uncertain ranges of soil depth.

Models with VPD and without Leaf Fraction (Eqs. \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51); followed by model with both VPD and Leaf Fraction (Eq. \@ref(eq:M4); Fig. S\@ref(fig:other-erd-mods)). Models with leaf fraction had only three species with sufficient, indvidual species-level goodness-of-fit (Eqs. \@ref(eq:M3 & eq:M4) versus seven for models without (Eqs. \@ref(eq:M1) & \@ref(eq:M2)). For rest of the analyses we therefore used species ERD estimates from model (\@ref(eq:M2)) as it was also a parsimonious model compared to \@ref(eq:M4) (Fig. \@ref(fig:erd-iso)).

@@

To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance  ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{TLP}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~Stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$).

@@

We calculated mortality rates for large trees (>=10 cm DBH) of xx canopy species for each of the seven census interval ($t$) in the 33 year record (1985-2015; $Mort_{t}$) and a species mean mortality rate across those intervals ($Mort_{mean} = \frac{1}{t}\sum_{t=1}^{6} {Mort_{t}}$). To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{80,Leaf}$.

@@

\
In this forest modelled Effective Rooting Depths (ERD) for the 32 large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)). For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{TLP}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~Stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).

```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for 32 large, canopy species of Barro Colorado Island. See Table S2 for scientific names of species codes")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
```

@@

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{80,Leaf}$ over 1990-2015, where z is the soil depth matching species ERD--for xx canopy species found on Barro Colorado Island. EAch circle resepresents a species. Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```
